Mixture Models for Single Cell Assays

### 2012-03-09 MIMOSA.tex
Updated multinomial dirichlet description
Added gradient and hessian for responder and non-responder components.

Signed-off-by: Greg Finak <gfinak@fhcrc.org>

For the null component, where $\bar{p}_{s}=\bar{p}_{u}$ the marginal likelihood is given by: 
\begin{align}
\mathrm{L}(\bar{n}_s,\bar{n}_u,N_s,N_u|\bar{\alpha}_u) &= \prod_{i=0}^P\frac{ \mathrm{B_j}(\bar{\alpha}_{u}+\bar{n}_{ui}+\bar{n}_{si})}{\mathrm{B_j}(\bar{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_{j=1}^J n_{sij}!} \cdot \frac{N_{ui}!}{\prod_{j=1}^J n_{uij}!}\\
\end{align}
Where $\mathrm{B_j}$ is the $\mathrm{j}$--dimensional Beta function: $\frac{\prod_{j=1}^J\Gamma(\alpha_j)}{\Gamma(\sum\alpha_j)}$.

@@

>>>>

For the null component, where $\bar{p}_{s}=\bar{p}_{u}$ the marginal likelihood is given by: 
\begin{align}
\mathrm{L}(\bar{n}_s,\bar{n}_u,N_s,N_u|\bar{\alpha}_u) &= \prod_{i=0}^P\frac{ \mathrm{B_j}(\bar{\alpha}_{u}+\bar{n}_{ui}+\bar{n}_{si})}{\mathrm{B_j}(\bar{\alpha}_u)} \cdot \frac{N_{si}!}{\prod_{j=1}^J n_{sij}!} \cdot \frac{N_{ui}!}{\prod_{j=1}^J n_{uij}!}\label{NRmd}
\end{align}
Where $\mathrm{B_j}$ is the $\mathrm{j}$--dimensional Beta function: $\frac{\prod_{j=1}^J\Gamma(\alpha_j)}{\Gamma(\sum\alpha_j)}$.

@@

The marginal likelihood of the non--responding component, given in eq~\eqref{NRmd} can be rewritten as:
\begin{align}
\begin{split}
l(\bar{\alpha}_u|D)=\sum_i \left[ \log \Gamma (\sum_j \alpha_{uj}) - \log \Gamma (\sum_j \alpha_{uj}+n_{uij}+n_{sij})+ \sum_j (\log \Gamma(\alpha_{uj}+n_{uij}+n_{sij}) \right.\\ \left. -\log\Gamma(\alpha_{uj}))+ \log\Gamma(N_{si}+1) +\log\Gamma(N_{ui}+1)-\right.\\ \left.\left[\sum_j(\log\Gamma(n_{sij}+1)+\log\Gamma(n_{uij}+1))\right]\right]\label{mdloglike}
\end{split}
\end{align}

The gradient is given by:
\begin{align}
\frac{\partial l}{ \partial \alpha_{uj}} = \sum_i \Psi(\sum_j \alpha_{uj}) - \Psi (\sum_j \alpha_{uj}+n_{uij}+n_{sij}) + \Psi(\alpha_{uj}+n_{uij}+n_{sij}) - \Psi(\alpha_{uj})\label{mdgrad}
\end{align}

and the Hessian is:
\begin{align}
j&\ne k; &\frac{\partial^2 l}{\partial \alpha_{uj}\alpha_{uk}} & = \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})\\
j&=k;&\frac{\partial^2l}{\partial \alpha_{uj}^2} &= \sum_i \Psi'(\sum_j\alpha_{uj})-\Psi'(\sum_j\alpha_{uj}+n_{uij}+n_{sij})+\Psi'(\alpha_{uj}+n_{uij}+n_{sij})-\Psi'(\alpha_{iuj})
\end{align}

The marginal likelihood of the responding component in eq~\ref{eq:postmult} can be rewritten as:
\begin{align}
\begin{split}
\sum_i\bigg[\sum_j \big[ \log\Gamma(\alpha_{uj}+n_{uij})+\log\Gamma(\alpha_{sj}+n_{sij}) -\log\Gamma(\alpha_{uj}) -\log\Gamma(\alpha_{sj})\\
-\log\Gamma(n_{sij}+1)-\log\Gamma(n_{uij}+1)\big]+\log\Gamma(\sum_j\alpha_{sj})+\log\Gamma(\sum_j\alpha_{uj})\\
-\log\Gamma(\sum_j\alpha_{uj}+n_{uij})-\log\Gamma(\sum_j\alpha_{sj}+n_{sij})+\log\Gamma(N_{si}+1)+\log\Gamma(N_{ui}+1)\bigg]
\end{split}
\end{align}
with gradient:
\begin{align}
\frac{\partial l}{\partial\alpha_{\phi j}} = \sum_i \Psi(\alpha_{\phi j}+n_{\phi ij})-\Psi(\alpha_{\phi j})+\Psi(\sum_j \alpha_{\phi j}) - \Psi(\sum_j \alpha_{\phi j}+n_{\phi ij}): (\phi \in {s,u})
\end{align}
and Hessian:
\begin{align}
\frac{\partial^2l}{\partial \alpha_{\phi j}\alpha_{\phi k}}& \overset {k=j}{=}\sum_i \Psi'(\alpha_{\phi j}) - \Psi'(\alpha_{\phi j}) + \Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})\\
\frac{\partial^2l}{\partial \alpha_{\phi j}^2}&=\Psi'(\sum_j \alpha_{\phi j}) - \Psi'(\sum_j\alpha_{\phi j}+n_{\phi ij})
\end{align}
